A Meta-analysis of Longevity Estimates of Mosquito Vectors of Disease: Supplementary Material

### 2018-11-23 MS.tex
First draft of article

Mosquitoes are responsible for more human deaths than any other
organism, yet we still know relatively little about their ecology.
Mosquito lifespan is a key determinant of transmission strength for the
diseases they vector, but the field experiments used to determine this
quantity -- mark-release-recapture (MRR) studies and wild-caught
dissection of female mosquitoes -- produce estimates with high
uncertainty. In this paper, we use Bayesian hierarchical models to
analyse a previously-published database of 232 MRR experiments and
compile then analyse a database of 131 dissection studies to produce the
first ever species- and genus-level estimates of mosquito lifespan. Due
to the assumptions required to analyse the field data, we term our
estimates lower bounds on lifespan (LBL). Notably, for the major African
malaria vector \emph{Anopheles gambiae s.l.}, we estimate LBLs of 4.5
days (mean estimate; 25\%-75\% CI: 3.8-5.1 days for unfed female
mosquitoes from the MRR analysis) and 9.5 days (mean estimate;
25\%-75\% CI: 5.2-11.0 days from the dissection analysis); and an LBL of
4.3 days (mean estimate; 25\%-75\%
CI: 3.6-4.8 days, only present in the MRR database) for the
predominantly East-African vector \emph{A. funestus s.l}. We estimate
LBLs of 7.0 days (mean estimate; 25\%-75\% CI: 4.5-8.5 days from the
MRR analysis) and 5.0 days (mean estimate; 25\%-75\% CI: 3.5-5.1 days
from the dissection analysis) for \emph{Aedes aegypti}; and 12.1 days
(median estimate; 25\%-75\% CI: 10.0-13.7 days from the MRR analysis)
for \emph{Ae.} \emph{albopictus} -- the predominant vectors of dengue
fever, chikungunya and Zika. Our estimates indicate that there is
significant variation in lifespan across species, with most variation
explained by differences between genera. In correspondence with
laboratory studies, we estimate that female mosquitoes outlive males by
0.9 days on average (mean estimate; 25\%-75\% CI: 0.3-1.6 days). We
fit models incorporating mosquito senescence to the data, which allows
us to assess evidence for age-dependent mortality in mosquitoes across
different species. We determine that 8 of 33 species included in the MRR
database indicated evidence for senescence, versus only 2 of 25 species
from the dissection database. Our analysis applies a common framework to
the analysis of databases of MRR and dissection-based experiments,
allowing us to produce robust estimates of lower bounds on lifespan. It
also enables us to critically appraise each field method, highlighting a need for alternative field methods for measuring this
important mosquito characteristic.

@@

Parameters were estimated using Bayesian techniques with relative
uninformative priors for $\kappa$ and the parameters of
$\lambda(t)$, but assuming a prior for $\psi$ indicating a low
recapture probability (bounded in part by knowledge of the maximum daily
recapture rates; see SOM). We used a Bayesian hierarchical model to estimate
distributions of lifespan at the species and the genus levels, and
across the complete data set. This procedure assumes that there is a
distribution of lifespan parameters for each species from which those
governing individual MRR time series are sampled, and similarly a
distribution at the genus level from which those for individual species
are derived (rather akin to random effects in classical statistics).
Within this framework we can also allow the parameters for individual
time series to be influenced by co-variates such as differences in
experimental methodology. As in the estimation of the parameters of the
individual experiments, relative uninformative priors were set for the
parameters of the hierarchical models except for $\psi$ where again a
distribution representing low recapture probabilities was assumed.
Posterior distributions were derived using Markov Chain Monte Carlo
(MCMC) methods with convergence assessed using the \(\hat{R}\) statistic
\citep{gelman1992inference}. The predictive power of the model was assessed
using \emph{K}-fold cross validation which tests the ability of the
model fitted to part of the data to predict the rest using multiple
different partitions. Further details of the prior specification,
fitting and validation through posterior predictive checks are given in
the SOM.

@@

In this study, we applied a Bayesian hierarchical framework to the
analysis of a database of mark-release-recapture experiments and a
database of mosquito dissection studies to estimate mosquito lifespan.
By applying a single framework, this allows us to effectively synthesise
information from the disparate experiments which, individually, estimate
lifespan with considerable uncertainty. Due to the assumptions required
to analyse the field data, our estimates represent lower bounds on
lifespan (LBL). Across both meta-analyses, the estimated LBLs were
mostly less than 10 days, hinting that only a small proportion of
mosquitoes may live long enough to transmit disease. We determined that
LBL varies across species and genera, although most variance is
explained by genus. The MRR analysis includes experiments conducted on
each sex individually, and we estimate that, on average, males live
shorter lives than females. Pre-release feeding with sugar also
lengthens lifespan across all three genera, although this effect is less
marked than the sex differences. In contrast to a number of lab-based
experiments, temperature was not determined to significantly impact
lifespan. By fitting a range of survival models to the data in both
meta-analyses, we could assess evidence for age-dependent mortality.
Overall, we conclude that the evidence is mixed: in the MRR experiments,
in 8 of 33 species we found evidence for mosquito senescence, whereas in
only 2 of 25 species included in the dissection analysis were better fit
by a model incorporating an increasing risk of mortality with age.

MRR experiments are known to produce downwardly-biased estimates of
lifespan. Lab experiments have demonstrated that marking can negatively
impact survival (Verhurst et al., 2013; Dickens and Brant, 2014),
resulting in artificially depressed survival. MRR studies typically
cannot differentiate between a mosquito dying and dispersal from the
study area meaning that lifespan will be underestimated. In this study,
we found a positive correlation between lifespan estimates and the
density of traps, indicating that better trapping coverage likely raises
estimates towards their real value. We conducted an \emph{in silico}
Monte Carlo study to determine how accurately we could estimate mosquito
lifespan given study parameters in an ideal MRR experiment, where the
assumptions of no emigration and harmless marking are fully satisfied
(see SOM for full details). This work indicated that for many of the
experiments, the short study lengths or typical numbers of mosquitoes
released, results in considerable uncertainty in lifespan estimates (Fig.
\ref{S-fig:mrr_mcPowerAnalysis}). This indicates that statistical power can be substantially
increased by pooling data across experiments as we did using a Bayesian
hierarchical model.

@@

proportion of ovarioles (Hoc and Wilkes, 1995), particularly in African
vector species (Gillies and Wilkes, 1965), the related issue of locating

@@

(Fox and Brust, 1994), and the variation in numbers of ovariolar
dilations for mosquitoes of the same, known, physiological age (Kay,
1979; Russell, 1986; Hugo et al., 2008). Indeed there is considerable

@@

term coined by Fox and Brust, 1994) headed by Polovodana (Polovodova,
1949) and Detinova (Detinova, 1962) considers dilations to result from
normal oogenesis, a `New School' headed by Lange and Hoc (Lange and Hoc,
1981) has challenged this assertion. The New School believe that only

@@

oogenesis destroys the sack-like structures (Fox and Brust, 1994). This

@@

older specimens (Fox and Brust, 1994), and likely biases lifespan

@@

mosquitoes by Hugo et al., 2008, who conclude that only a small

@@

gonotrophic cycles: MRR studies (see, for example, Gillies and Wilkes,
1965), where marked mosquitoes are recaptured and dissected to determine

@@

females, or their progeny (see, for example, Afrane et al., 2005).

@@

results in a relative paucity (Gillies and Wilkes, 1965) or abundance
(Clements and Paterson, 1981) of nulliparous individuals. In our

@@

traps (Detinova, 1962; Gillies and Wilkes, 1965). We chose to not

@@

(Detinova, 1962), based on dichotomous categorisation of female mosquito

@@

The different nature of the assumptions of each of the two methods means
they offer complimentary information on mosquito survival. We also note
that Polovodova's dissection-based studies require specialised expertise
which will often be unavailable, whereas MRR methods can more readily be
used. Furthermore, most if not all dissection methods that have been
used previously are only applicable to female mosquitoes, whereas MRR
can be applied to either sex and can additionally be used to determine
other ecological parameters (for example, population size and
dispersal). Although dissection data gives detailed of age-structure, we
thus foresee a continued reliance on MRR experiments in field
entomological experiments. Efforts to use both approaches concurrently
will be particularly useful and will allow quantification of the biases
induced by the assumptions of each. Similarly, MRR experiments releasing
large numbers of marked mosquitoes and recording
spatiotemporally-disaggregated captures of wild and re-caught marked
mosquitoes will be especially useful to estimate lifespan and dispersal.

To compare estimates of lifespan derived from MRR with those from
dissection-based methods, we display the estimates of lifespan from
those ten species occurring in both databases in a single plot (Fig. \ref{fig:comparison}). In is reassuring that there is correlation between estimates from both approaches, although the small sample size likely hindered our ability
to determine statistical significance. In both cases, we estimate that
\emph{A. sergentii} was amongst the longest lived of the anopheline
species with a posterior median LBL of 8.7 days (median estimate;
25\%-75\% CI: 5.9-13.8 days) from the MRR analysis and 10.0 days (median
estimate; 25\%-50\% CI: 7.6-14.0 days) from the analysis of dissection
studies. This species is a major vector of malaria in the Sahara (Sinka
et al., 2010), where to act as a disease vector it must persevere
through these hard conditions. It is reasonable to hypothesise that this
species should live longer than those in environments where the
potential for blood-feeding and oviposition is greater. The species with
the greatest discrepancy in the estimates was the major African malaria
vector \emph{A. gambiae s.l.}, where we estimated LBLs of 4.4 days
(median estimate; 25\%-75\% CI: 3.8-5.1 days for unfed female) from the
MRR analysis and 7.6 days (median estimate; 25\%-75\% CI: 5.2-11.0) from
the dissection analysis. Across genera, the greatest discrepancy in
estimates was for \emph{Culex} (a posterior median of 2.5 days from the
MRR versus 4.4 days from the dissection analysis) with the smallest
discrepancy for \emph{Anopheles} (4.8 versus 5.0 days). By contrast for
\emph{Aedes} the estimates from the MRR studies (6.9 days) exceed those
of dissection-based studies (3.5 days). Across all studies we estimate
from the MRR analysis that mean mosquito lifespan is 4.6 days versus 5.1
days from the dissection-based studies. Some of the differences in these
group-level estimates between the two approaches is likely due to
environmental and genetic differences between mosquitoes in the
experiments that were analysed in each meta-analysis. However, we
believe that part of the discrepancy can be explained by the
methodological differences in approaches. We speculate that differences
in dispersal rate can explain some of the discrepancy. Both
\emph{Anopheles} and \emph{Culex} mosquitoes are generally thought to
fly farther during their lifetimes than \emph{Aedes} {[}Charles, do you
have a reference here?{]}, meaning that the estimates from MRR-based
approaches will be most downwardly-biased for these genera. This is
supported by our results since the dissection-based estimates
(themselves not reliant on assumptions about dispersal) exceed the MRR
estimates for \emph{Anopheles} and \emph{Culex} mosquitoes, but not for
\emph{Aedes}.

It is widely believed mosquitoes live artificially long under the benign
conditions of the laboratory. We find it informative to consider
estimates of lifespan derived from observations of such populations as
they constitute an upper bound on the lifespan of wild populations.
Also, since the numbers of mosquitoes involved in large cage experiments
often numbers in the thousands, these estimates have lower uncertainty
than those from field experiments although are typically conducted on
highly inbred mosquito strains. Styer et al. (2007), using colonies of
45,054 female and 55,997 male \emph{Ae. aegypti}, determined that
females lived nearly twice as long as males; the median lifespan was
estimated as 31.69 $\pm$ 0.06 days for females and 16.39 $\pm$ 0.03 days
for males. A similar study by Dawes et al., 2009 with a lab colony of
over 1000 female \emph{A. stephensi} found similar estimates for median
lifespan (31-42 days). These estimates are many multiples of the average
estimates that result from our analysis of field data which, as
discussed, represent lower bound estimates. Without an unbiased method
to measure mosquito lifespan, however, it is difficult to quantify and
explain the gap that exists between field and laboratory lifespans. The
development of additional methods to estimate mosquito age, such as
`Near-Infrared Spectroscopy' (Mayagaya et al., 2009; Sikulu et al.,
2011; Lambert et al., 2018), if they are proven to work in the field,
may be of considerable worth here.

We conducted a power analysis of MRR experiments to determine
whether typical experimental characteristics could detect senescence.
Here we calculated the power of a maximum likelihood estimator of the
`senescence parameter' $\beta$ of the Gompertz survival function (see Table
S\#) for case study populations with three different levels of
senescence (Fig. \ref{S-fig:mrr_mcPowerAnalysis_senescence}A). This analysis indicated that power to detect
senescence strongly depends on study length (Fig. \ref{S-fig:mrr_mcPowerAnalysis_senescence}B) but is
insensitive to release size (Fig. \ref{S-fig:mrr_mcPowerAnalysis_senescence}C). Clements and Patterson (1981)
conducted a meta-analysis of MRR and dissection-based field experiments
and found evidence of an increasing risk of mortality hazard with age
that is similar in magnitude to that of the `mild' case considered
above. For this case, detecting senescence with a power of 80\% requires
a study length of at least 18 days. Since the median study length for
experiments included in our analysis was 10 days (Table S\#) this could
partly explain our failure to detect senescence at the species level. A
number of experiments have found evidence of age-dependence in
laboratory populations (Styer et al., 2007; Dawes et al., 2009).
However, the artificially benign environment of the laboratory means
mosquitoes live considerably longer than in the wild, where they may die
because of exogenous factors, before the effects of physiological
decline have had time to manifest. Field experiments have also found
evidence for age-dependent mortality. Harrington et al. (2008) conducted
a field experiment where mosquitoes reared under laboratory conditions
were marked and released at different ages. Analysis of the resultant
MRR time-series indicated that mosquito mortality increases with age at
release. It is possible, however, that this field experiment suffers
from the same biases as laboratory-based approaches, because the
released mosquitoes were often of ages considerably higher (up to 20
days) than typical estimates of wild mosquito lifespan.

As ethical concerns of contributing to disease burden are more often
considered, it is now less common for MRR experiments to release female
mosquitoes versus males than historically (Fig. \ref{S-fig:mrr_sexReleasesOverTime}). Our analysis
indicates that females outlive male mosquitoes by approximately 0.9 days
(Fig. \ref{fig:mrr_sexDifferences_without_sugar_nor_blood}), meaning that differences between the sexes may exist for
other ecological parameters determinable by MRR. This suggests that
continued field entomological work on contained releases of mosquitoes
in semi-field sites or large microcosms may be a valuable source of
information on female mosquitoes.

Our estimates of LBL indicate that mosquitoes that were sugar-fed prior
to release lived on average 0.7 days longer than those that were unfed
(, suggesting the potential value of this underappreciated aspect of the
mosquito ecology to the insects. It may also partly explain the recent
successes in the use of Attractive Toxic Sugar Baits (ATSB) as a vector
control intervention (Schlein and Muller references). More research is
needed, however, to identify the sugar-feeding frequency and food
sources for wild populations.

There is considerable evidence from the laboratory that temperature
modulates mosquito ecology and behaviour (Matt Thomas references). The
locations and times of year over which the MRR studies were conducted
encompassed a large range of average air temperatures, from
approximately 10 \textsuperscript{o}C to 35 \textsuperscript{o}C and,
within this, we determined no relationship between lifespan and
temperature across all time series (Fig. \ref{fig:mrr_temperature}) or, for any of the species
with the most data (Fig. \ref{S-fig:mrr_ThreeSpeciesVersusTemperature}). It is possible that by considering a raw
average of air temperature across the month, this ignored, more complex, interactions between temperature and lifespan. It is also possible that by ignoring the effects of
rainfall (the historical data on rainfall is less likely to be reliable
for a given location), that this masked a more complex interaction
between longevity and temperature. The observed laboratory relationship
between lifespan and temperature, however, may not be as robust in the
field if mosquitoes adjust their behaviours (such as, by seeking shade)
in reaction to changes in temperature. More work exploring the
relationship between mosquito ecology and temperature in semi-field
experiments may be useful in probing these interactions further.

In this work, we have used modern statistical methods to synthesise precious field data conducted by entomologists past and present, to produce lower bound estimates of mosquito lifespan. The importance of vector mortality for disease transmission has long been recognised, however,
since even before 1957, when George Macdonald formulated the
now famous Ross-Macdonald equation of R\textsubscript{0} for malaria. Indeed, the recent declines in malaria prevalence in Sub-Saharan Africa were likely due to upscaling of interventions (insecticide-treated bednets and indoor residual spraying) that aim to reduce mosquito lifespan (Bhatt et al., 2015). Worryingly, resistance to pyrethroids, the only class of insecticide used in current insecticide-treated bednets and likely the only product to come to market in the near
future, has been determined to be widespread and increasing in intensity
across Sub-Saharan Africa (WHO reference 2018). This alarming trend
highlights the need for continued MRR and dissection-based studies to
monitor the effectiveness of bednets and determine whether more
expensive alternatives, such as nets incorporating piperonyl butoxide be
deployed. It also emphasises the need for investment in new tools for
real time monitoring of mosquito populations. In recent years, there has been considerable funding allocated to molecular and genomic research
into mosquitoes that strengthens existing interventions and suggest
novel control strategies. Without commensurate funding allocated to
applied vector ecology, our lack of knowledge in this area threatens our
opportunity to capitalise on molecular advances and potentially hinders
our ability to control of mosquito-borne disease.

@@

>>>>

Mosquitoes are responsible for more human deaths than any other
organism, yet we still know relatively little about their ecology.
Mosquito lifespan is a key determinant of transmission strength for the
diseases they vector, but the field experiments used to determine this
quantity -- mark-release-recapture (MRR) studies and wild-caught
dissection of female mosquitoes -- produce estimates with high
uncertainty. In this paper, we use Bayesian hierarchical models to
analyse a previously-published database of 232 MRR experiments and
compile then analyse a database of 131 dissection studies to produce the
first ever species- and genus-level estimates of mosquito lifespan. Due
to the assumptions required to analyse the field data, we term our
estimates lower bounds on lifespan (LBL). Notably, for the major African
malaria vector \emph{Anopheles gambiae s.l.}, we estimate LBLs of 4.5
days (mean estimate; 25\%-75\% CI: 3.8-5.1 days for unfed female
mosquitoes from the MRR analysis) and 9.5 days (mean estimate;
25\%-75\% CI: 5.2-11.0 days from the dissection analysis); and an LBL of
4.3 days (mean estimate; 25\%-75\%
CI: 3.6-4.8 days, only present in the MRR database) for the
predominantly East-African vector \emph{A. funestus s.l}. We estimate
LBLs of 7.0 days (mean estimate; 25\%-75\% CI: 4.5-8.5 days from the
MRR analysis) and 5.0 days (mean estimate; 25\%-75\% CI: 3.5-5.1 days
from the dissection analysis) for \emph{Aedes aegypti}; and 12.1 days
(mean estimate; 25\%-75\% CI: 10.0-13.7 days from the MRR analysis)
for \emph{Ae.} \emph{albopictus} -- the predominant vectors of dengue
fever, chikungunya and Zika. Our estimates indicate that there is
significant variation in lifespan across species, with most variation
explained by differences between genera. In correspondence with
laboratory studies, we estimate that female mosquitoes outlive males by
0.9 days on average (mean estimate; 25\%-75\% CI: 0.3-1.6 days). We
fit models incorporating mosquito senescence to the data, which allows
us to assess evidence for age-dependent mortality in mosquitoes across
different species. We determine that 8 of 33 species included in the MRR
database indicated evidence for senescence, versus only 2 of 25 species
from the dissection database. Our analysis applies a common framework to
the analysis of databases of MRR and dissection-based experiments,
allowing us to produce robust estimates of lower bounds on lifespan. It
also enables us to critically appraise each field method, highlighting a need for alternative field methods for measuring this
important mosquito characteristic.

@@

Parameters were estimated using Bayesian techniques with relative
uninformative priors for $\kappa$ and the parameters of
$\lambda(t)$, but assuming a prior for $\psi$ indicating a low
recapture probability (bounded in part by knowledge of the maximum daily
recapture rates; see SOM). We used a Bayesian hierarchical model to estimate
distributions of lifespan at the species and the genus levels, and
across the complete data set. This procedure assumes that there is a
distribution of lifespan parameters for each species from which those
governing individual MRR time series are sampled, and similarly a
distribution at the genus level from which those for individual species
are derived (rather akin to random effects in classical statistics).
Within this framework we can also allow the parameters for individual
time series to be influenced by co-variates such as differences in
experimental methodology. As in the estimation of the parameters of the
individual experiments, relative uninformative priors were set for the
parameters of the hierarchical models except for $\psi$ where again a
distribution representing low recapture probabilities was assumed.
Posterior distributions were derived using Markov Chain Monte Carlo
(MCMC) methods with convergence assessed using the \(\hat{R}\) statistic
\citep{gelman1992inference}. The predictive power of the model was assessed
using \emph{K}-fold cross validation which tests the ability of the
model fitted to part of the data to predict the rest using multiple
different partitions. Further details of the prior specification,
fitting and validation through posterior predictive checks \citep{lambert2018student} are given in
the SOM.

@@

In this study, we applied a Bayesian hierarchical framework to the
analysis of a database of mark-release-recapture experiments and another for mosquito dissection studies to estimate mosquito lifespan.
By applying a single framework, this allows us to effectively synthesise
information from the disparate experiments which, individually, estimate
lifespan with considerable uncertainty. Due to the assumptions required
to analyse the field data, our estimates represent lower bounds on
lifespan (LBL). Across both meta-analyses, the estimated LBLs were
mostly less than 10 days, hinting that only a small proportion of
mosquitoes may live long enough to transmit disease. We determined that
LBL varies across species and genera, although most variance is
explained by genus. The MRR analysis includes experiments conducted on
each sex individually, and we estimate that, on average, males live
shorter lives than females. Pre-release feeding with sugar also
lengthens lifespan across all three genera, although this effect is less
marked than the sex differences. In contrast to a number of lab-based
experiments \citep{yang2009assessing,brady2013modelling}, temperature was not determined to significantly impact
lifespan. By fitting a range of survival models to the data in both
meta-analyses, we could assess evidence for age-dependent mortality.
Overall, we conclude that the evidence is mixed: in the MRR experiments,
in 8 of 33 species we found evidence for mosquito senescence, whereas in
only 2 of 25 species included in the dissection analysis were better fit
by a model incorporating an increasing risk of mortality with age.

MRR experiments are known to produce downwardly-biased estimates of
lifespan. Lab experiments have demonstrated that marking can negatively
impact survival \citep{verhulst2013advances,dickens2014effects}
resulting in artificially depressed survival. MRR studies typically
cannot differentiate between a mosquito dying and dispersal from the
study area meaning that lifespan will be underestimated. In this study,
we found a positive correlation between lifespan estimates and the
density of traps, indicating that better trapping coverage likely raises
estimates towards their real value. We conducted an \emph{in silico}
Monte Carlo study to determine how accurately we could estimate mosquito
lifespan given study parameters in an ideal MRR experiment, where the
assumptions of no emigration and harmless marking are fully satisfied
(see SOM for full details). This work indicated that for many of the
experiments, the short study lengths or typical numbers of mosquitoes
released, results in considerable uncertainty in lifespan estimates (Fig.
\ref{S-fig:mrr_mcPowerAnalysis}). This indicates that statistical power can be substantially
increased by pooling data across experiments as we did using a Bayesian
hierarchical model.

@@

proportion of ovarioles \citep{hoc1995ovariole}, particularly in African
vector species \citep{gillies1965study}, the related issue of locating

@@

\citep{fox1994dilatations}, and the variation in numbers of ovariolar
dilations for mosquitoes of the same, known, physiological age \citep{kay1979age,russell1986population,hugo2008evaluations}. Indeed there is considerable

@@

term coined by Fox and Brust, 1994) headed by Polovodana \citep{polovodova1949determination} and Detinova \citep{detinova1962age} considers dilations to result from
normal oogenesis, a `New School' headed by Lange and Hoc \citep{lange1981abortive} has challenged this assertion. The New School believe that only

@@

oogenesis destroys the sack-like structures \citep{fox1994dilatations}. This

@@

older specimens \citep{fox1994dilatations}, and likely biases lifespan

@@

mosquitoes by Hugo et al. (2008), who conclude that only a small

@@

gonotrophic cycles: MRR studies (see, for example, \cite{gillies1965study}), where marked mosquitoes are recaptured and dissected to determine

@@

females, or their progeny (see, for example, \cite{afrane2005effects}).

@@

results in a relative paucity \citep{gillies1965study} or abundance
\citep{clements1981analysis} of nulliparous individuals. In our

@@

traps \citep{detinova1962age,clements1981analysis}. We chose to not

@@

\cite{detinova1962age}, based on dichotomous categorisation of female mosquito

@@

The different nature of the assumptions of each of the two methods means
they offer complimentary information on mosquito survival. We also note
that Polovodova's dissection-based studies require specialised expertise
which will often be unavailable, whereas MRR methods can more readily be
used. Furthermore, most if not all dissection methods that have been
used previously are only applicable to female mosquitoes, whereas MRR
can be applied to either sex and can additionally be used to determine
other ecological parameters (for example, population size and
dispersal). Although dissection data gives detailed of age-structure, we
thus foresee a continued reliance on MRR experiments in field
entomological experiments. Efforts to use both approaches concurrently
will be particularly useful and will allow quantification of the biases
induced by the assumptions of each. Similarly, MRR experiments releasing
large numbers of marked mosquitoes and recording
spatiotemporally-disaggregated captures of wild and re-caught marked
mosquitoes will continue be useful in estimating lifespan and dispersal.

To compare estimates of lifespan derived from MRR with those from
dissection-based methods, we display the estimates of lifespan from
those ten species occurring in both databases in a single plot (Fig. \ref{fig:comparison}). In is reassuring that there is correlation between estimates from both approaches, although the small sample size likely hindered our ability
to determine statistical significance. In both cases, we estimate that
\emph{A. sergentii} was amongst the longest lived of the anopheline
species with a posterior median LBL of 12.4 days (mean estimate;
25\%-75\% CI: 5.9-13.8 days) from the MRR analysis and 11.9 days (mean
estimate; 25\%-50\% CI: 7.6-14.0 days) from the analysis of dissection
studies. This species is a major vector of malaria in the Sahara \citep{sinka2010dominant}, where to act as a disease vector it must persevere
through these hard conditions. It is reasonable to hypothesise that this
species should live longer than those in environments where the
potential for blood-feeding and oviposition is greater. The species with
the greatest discrepancy in the estimates was the major African malaria
vector \emph{A. gambiae s.l.}, where we estimated LBLs of 4.5 days
(mean estimate; 25\%-75\% CI: 3.8-5.1 days for unfed female) from the
MRR analysis and 9.5 days (mean estimate; 25\%-75\% CI: 5.2-11.0) from
the dissection analysis. Across genera, the greatest discrepancy in estimates was for
\emph{Aedes}, where the estimates from the MRR studies (8.1 days) are considerably longer than those of dissection-based studies (3.5 days).
This was followed by \emph{Culex} (a posterior mean of 2.9 days from the
MRR versus 4.9 days from the dissection analysis) with the smallest
discrepancy for \emph{Anopheles} (6.8 versus 6.4 days). Across all studies we estimate
from the MRR analysis that mean mosquito lifespan is 6.0 days versus 5.5
days from the dissection-based studies. Some of the differences in these
group-level estimates between the two approaches is likely due to
environmental and genetic differences between mosquitoes in the
experiments that were analysed in each meta-analysis. However, we
believe that part of the discrepancy can be explained by the
methodological differences in approaches. We speculate that differences
in dispersal rate can explain some of the discrepancy. Both
\emph{Anopheles} and \emph{Culex} mosquitoes are generally thought to
fly farther during their lifetimes than \emph{Aedes} {[}Charles, do you
have a reference here?{]}, meaning that the estimates from MRR-based
approaches will be most downwardly-biased for these genera. This is
supported by our results since the dissection-based estimates
(themselves not reliant on assumptions about dispersal) are similar or exceed the MRR
estimates for \emph{Anopheles} and \emph{Culex} mosquitoes, but not for
\emph{Aedes}.

It is widely believed mosquitoes live artificially long under the benign
conditions of the laboratory. We find it informative to consider
estimates of lifespan derived from observations of such populations as
they constitute an upper bound on the lifespan of wild populations.
Also, since the numbers of mosquitoes involved in large cage experiments
often numbers in the thousands, these estimates have lower uncertainty
than those from field experiments although are typically conducted on
highly inbred mosquito strains. \cite{styer2007mosquitoes}, using colonies of
45,054 female and 55,997 male \emph{Ae. aegypti}, determined that
females lived nearly twice as long as males; the median lifespan was
estimated as 31.69 $\pm$ 0.06 days for females and 16.39 $\pm$ 0.03 days
for males. A similar study by \cite{dawes2009anopheles} with a lab colony of
over 1000 female \emph{A. stephensi} found similar estimates for median
lifespan (31-42 days). These estimates are many multiples of the average
estimates that result from our analysis of field data which, as
discussed, represent lower bound estimates. Without an unbiased method
to measure mosquito lifespan, however, it is difficult to quantify and
explain the gap that exists between field and laboratory lifespans. The
development of additional methods to estimate mosquito age, such as
`Near-Infrared Spectroscopy' \citep{mayagaya2009non,sikulu2011evaluating,lambert2018monitoring} if they are proven to work in the field,
may be of considerable worth here.

We conducted a power analysis of MRR experiments to determine
whether typical experimental characteristics could detect senescence.
Here we calculated the power of a maximum likelihood estimator of the
`senescence parameter' $\beta$ of the Gompertz survival function (see Table
S\ref{S-tab:mrr_survivalDescription}) for case study populations with three different levels of
senescence (Fig. \ref{S-fig:mrr_mcPowerAnalysis_senescence}A). This analysis indicated that power to detect
senescence strongly depends on study length (Fig. \ref{S-fig:mrr_mcPowerAnalysis_senescence}B) but is
insensitive to release size (Fig. \ref{S-fig:mrr_mcPowerAnalysis_senescence}C). Clements and Patterson (1981)
conducted a meta-analysis of MRR and dissection-based field experiments
and found evidence of an increasing risk of mortality hazard with age
that is similar in magnitude to that of the `mild' case considered
above. For this case, detecting senescence with a power of 80\% requires
a study length of at least 18 days. Since the median study duration for
experiments included in our analysis was 10 days (Table S\ref{S-tab:mrr_IndividualData}) this could
partly explain our failure to detect senescence at the species level. A
number of experiments have found evidence of age-dependence in
laboratory populations \citep{styer2007mosquitoes,dawes2009anopheles}.
However, the artificially benign environment of the laboratory means
mosquitoes live considerably longer than in the wild, where they may die
because of exogenous factors, before the effects of physiological
decline have had time to manifest. Field experiments have also found
evidence for age-dependent mortality. Harrington et al. (2008) conducted
a field experiment where mosquitoes reared under laboratory conditions
were marked and released at different ages. Analysis of the resultant
MRR time-series indicated that mosquito mortality increases with age at
release. It is possible, however, that this field experiment suffers
from the same biases as laboratory-based approaches, because the
released mosquitoes were often of ages considerably higher (up to 20
days) than typical estimates of wild mosquito lifespan.

As ethical concerns of contributing to disease burden are more often
considered, it is now less common for MRR experiments to release female
mosquitoes versus males than historically (Fig. \ref{S-fig:mrr_sexReleasesOverTime}). Our analysis
indicates that females outlive male mosquitoes by approximately 0.9 days
(Fig. \ref{fig:mrr_sexDifferences_without_sugar_nor_blood}), meaning that differences between the sexes may exist for
other ecological parameters determinable by MRR. This suggests that
continued field entomological work on contained releases of mosquitoes
in semi-field sites or large microcosms may be a valuable source of
information on female mosquito ecology.

Our estimates of LBL indicate that mosquitoes that were sugar-fed prior
to release lived on average 0.7 days longer than those that were unfed (Fig. \ref{S-fig:mrr_female_blood_sugar})
suggesting the potential value of this underappreciated aspect of the
mosquito ecology to the insects. It may also partly explain the recent
successes in the use of Attractive Toxic Sugar Baits as a vector
control intervention \citep{muller2008decline,muller2010effective,muller2010field,muller2010successful,beier2012attractive}. More research is
needed, however, to identify the sugar-feeding frequency and food
sources for wild populations.

There is evidence mainly from laboratory studies that temperature
modulates mosquito ecology and behaviour \citep{yang2009assessing,brady2013modelling,murdock2012complex,beck2013effect}. The
locations and times of year over which the MRR studies were conducted
encompassed a large range of average air temperatures, from
approximately 10 \textsuperscript{o}C to 35 \textsuperscript{o}C and,
within this, we determined no relationship between lifespan and
temperature across all time series (Fig. \ref{fig:mrr_temperature}) or, for any of the species
with the most data (Fig. \ref{S-fig:mrr_ThreeSpeciesVersusTemperature}). It is possible that by considering a raw
average of air temperature across the month, this ignored, more complex, interactions between temperature and lifespan. It is also possible that by ignoring the effects of
rainfall (the historical data on rainfall is less likely to be reliable
for a given location), that this masked a more complex interaction
between longevity and temperature. The observed laboratory relationship
between lifespan and temperature, however, may not be as robust in the
field if mosquitoes adjust their behaviours (such as, by seeking shade)
in reaction to changes in temperature. More work exploring the
relationship between mosquito ecology and temperature in semi-field
experiments may be useful in probing these interactions further.

In this work, we have used modern statistical methods to synthesise precious field data conducted by entomologists past and present, to produce lower bound estimates of mosquito lifespan. The importance of vector mortality for disease transmission has long been recognised, however,
since even before 1957, when George Macdonald formulated the
now famous Ross-Macdonald equation of R\textsubscript{0} for malaria. Indeed, the recent declines in malaria prevalence in Sub-Saharan Africa were likely due to upscaling of interventions (insecticide-treated bednets and indoor residual spraying) that aim to reduce mosquito lifespan \citep{bhatt2015effect}. Worryingly, resistance to pyrethroids, the only class of insecticide used in current insecticide-treated bednets and likely the only product to come to market in the near
future, has been determined to be widespread and increasing in intensity
across Sub-Saharan Africa \citep{world2018global}. This alarming trend
highlights the need for continued MRR and dissection-based studies to
monitor the effectiveness of bednets and determine whether more
expensive alternatives, such as nets incorporating piperonyl butoxide be
deployed. It also emphasises the need for investment in new tools for
real time monitoring of mosquito populations. In recent years, considerable funding has been allocated to molecular and genomic research
into mosquitoes that strengthens existing interventions and suggest
novel control strategies. Without commensurate funding allocated to
applied vector ecology, our lack of knowledge in this area threatens our
opportunity to capitalise on molecular advances and potentially hinders
our ability to control of mosquito-borne disease.

@@

\printbibliography
